# Extensions

## Extensions 

### More bases

There are many options available in <span class="pack">mgcv</span> beyond cubic splines, including thin plate, duchon, p-splines and more, as well as ones for cyclical effects, geographical, and beyond.  See `?smooth.terms` for more information.

### Interactions

We can implement interactions of multiple smooth terms. The following shows a model in which we could combine the effects of year and budget via the <span class="func">te</span> function, along with the corresponding predictive plots.  Note that the 3d is interactive, so twist it however you like to see what's going on.

```{r gam2d, fig.show='hide', echo=1}
gam_model_2d = gam(rating ~  te(year, budget_2016),
                      data=movies_yearly)
pd = visreg::visreg2d(gam_model_2d, 'year', 'budget_2016')[1:3] # for some reason 2d doesn't have a plot= argument
names(pd) = c('Year', 'Budget', 'Rating')
pd$Rating = t(pd$Rating)
# because of a lack of  sense, x is y in plotting 
```

```{r contourheat, echo=FALSE}
colfun = rev(colorRampPalette(RColorBrewer::brewer.pal(9, 'RdBu'))(100))
with(lapply(pd, round, 2),
     plot_ly(x=~Year, y=~Budget,  z=Rating, type='heatmap', colors=colfun, width='75%'))  %>%
  layout(paper_bgcolor='rgba(0,0,0,0)', plot_bgcolor='rgba(0,0,0,0)')
with(lapply(pd, round, 2), plot_ly(x=Year, y=Budget,  z = Rating, type='surface', hoverinfo='none', colors=colfun, width='75%',
                                   lighting=list(roughness=.01, ambient=.9), 
                                   contours=list(x=list(highlightcolor=toRGB('gray90'), width=1),
                                                 y=list(highlightcolor=toRGB('gray90'), width=1),
                                                 z=list(highlightcolor=toRGB('gray90'), width=1))))  %>% 
  layout(paper_bgcolor='rgba(0,0,0,0)', plot_bgcolor='rgba(0,0,0,0)')
```

While we see the general year effect where the earlier and more recent movies are associated with higher ratings, it would appear that for more recent times, lower ratings are associated with bigger budget movies.

There is also a way to do this interaction in a manner similar to ANOVA decomposition with both main effects and interaction.  See the <span class="func">ti</span> function.


### Grouped Effects

We can apply a smooth for a continuous covariate across levels of some grouping factor.  In the following, we'll let year have a different effect for Action movies vs. other movies.

```{r gamby, echo=1}
gam_model_by = gam(rating ~  s(year, by=Action) + Action + budget_2016 + votes + length, 
                   data=movies_yearly_action)
pd = visreg::visreg(gam_model_by, xvar='year', by='Action', plot=F)$fit
pd %>% 
  rename(fit = visregFit) %>% 
  plot_ly() %>% 
  add_lines(~year, ~fit, color=~Action) %>% 
  theme_plotly()
```

In this case Action movies are generally rated worse, and suffered a second and sharper dip starting in the 80s. Also, their recent rise is not quite as sharp.  Another alternative specifies a specific type of smooth that would be similar to random slopes in a mixed model. We now turn to the mixed model approach with GAMs.


### Mixed Models

Consider the matrix $Z$ that includes the basis functions, with associated coefficients $\gamma$, we could then write our model in matrix form as
follows:

$$ y = X\beta + Z\gamma + \epsilon$$

If you've spent much time with mixed models, that probably looks familiar to you. This penalized regression approach not only looks like the mixed model, we can exploit this to produce the same results.  The <span class="pack">mgcv</span> package has many tools for mixed models.

For this example I will use the sleep study data that comes with the lme4 package. It has reaction times for some tests where 18 subjects were restricted to 3 hours of sleep each night for 10 days. We fit a model with (uncorrelated) random intercepts and random slopes for Days.

```{r gamm, echo=1:8}
library(lme4)
mod_lme4 = lmer(Reaction ~ Days + (1|Subject) + (0 + Days|Subject), sleepstudy)
mod_gamm = gam(Reaction ~ Days + s(Subject, bs='re') + s(Days, Subject, bs='re'), 
               data=sleepstudy, method='REML')

# summary(mod_lme4)
# summary(mod_gamm)
VarCorr(mod_lme4)
gam.vcomp(mod_gamm)
cbind(ranef(mod_lme4)$Subject,
      Int_gamm = coef(mod_gamm)[3:20],
      Days_gamm = coef(mod_gamm)[21:length(coef(mod_gamm))]) %>% 
  round(2) %>% 
  rename(Int_mixed = `(Intercept)`,
         Days_mixed = Days) %>% 
  select(Int_mixed, Int_gamm, Days_mixed, Days_gamm) %>% 
  DT::datatable(options=list(dom='pt'))
```

The results are essentially identical. The <span class="pack">mgcv</span> package can also use <span class="pack">lme4</span> and <span class="pack">nlme</span> under the hood via other functions (<span class="pack">gamm4</span> package and <span class="func">gamm</span> function respectively). Extending GAMs into the mixed model world opens the door to some very powerful modeling possibilities.  


### Spatial models

Consider a data set with latitude and longitude coordinates among other covariates used to model some target variable.  A spatial regression analysis might incorporate an approach to incorporate spatial covariance among the observation points (e.g. <span class="emph">kriging</span>).  Such an approach is a special case of <span class="emph">Gaussian process</span>.  Know what else happens to be a special case of a Gaussian process? That's right, a GAM!  As such we can add spatial models to the sorts of models covered by GAMs too.  The following gives a sense of the syntax.  The mgcv package even has a `'gp'` smooth, because why wouldn't it?

 
```{r gamspatpoint, eval=F}
gam(y ~ s(lat, lon, bs='gp'))  # Matern
```

What about the discrete case, where the spatial *random effect* is based on geographical regions?  This involves a penalty that is based on the adjacency matrix of the regions, where if there are $g$ regions the adjacency matrix is a $g \times g$ indicator matrix where 1 indicates region i is connected to region j, and 0 otherwise.  In addition an approach similar to that for a random effect is used to incorporate observations belonging to specific regions.  You'll be shocked to know that the mgcv package has a smooth option for that, `bs='mrf'`, where `mrf` stands for <span class="emph">Markov random field</span>, which is an undirected graph.  Plotting the smooth term is akin to displaying the spatial random effect.

The following shows example code. See the help for `mrf` for more detail.

```{r gamspatdiscrete, eval=F}
gam(crime ~ s(region, bs = "mrf", xt = polygonListObject), data = df, method = "REML")
```


### My god, it's full of STARs!

The incorporation of additive, mixed, and spatial regression models gets us to STARs or, <span class="emph">structured additive regression</span> models.  This is an extremely general and powerful modeling tool that can take you very, very far.

## Connections & Generalizations

### Adaptive Basis Function Models

GAMs can be seen as belonging to a family of what are called <span class="emph">adaptive basis function models</span>.  These include random forests, neural nets, and boosting approaches.  GAMs might be seen as a midpoint lying between highly interpretable standard linear models and the more black box methods just mentioned.


### Beyond GLM family distributions

The mgcv package provides a plethora of distributions to use in modeling that might useful including:

- tweedie
- negative binomial
- ordered categorical
- beta
- t
- zero-inflated poisson (ZIP)
- cox proportional hazards
- gaulss Gaussian model for both mean (location) and standard deviation (scale)
- gevlss same for generalized extreme value
- ziplss for a hurdle approach to zero-inflation
- multivariate normal
- multinomial logistic

### Random Effects

In addition to distribution extensions that could also be incorporated with random effects, providing modeling capabilities not generally offered in mixed model packages, mgcv also allows for correlated residuals structure as with nlme, but for the generalized case.  This is especially of interest in the longitudinal setting, where one expects residuals for observations closer in time to be more correlated.

### Gaussian Processes

Where the Gaussian distribution is over vectors and defined by a mean vector and covariance matrix, a <span class="emph">Gaussian Process</span> is a *distribution over functions*. A function $f$ is distributed as a Gaussian Process defined by a mean function $m$ and covariance function $k$. They have a close tie to RKHS methods, and generalize commonly used models in spatial modeling (kriging).  In addition, a GP can be interpreted as a standard multilayer perceptron neural net with a single hidden layer consisting of an infinite number of nodes.

$$f\sim \mathcal{GP}(m,k)$$

It turns out that GAMs with a tensor product or cubic spline smooth are maximum a posteriori (MAP) estimates of Gaussian processes with specific covariance functions and a zero mean function. In that sense one might segue nicely to Gaussian processes if familiar with additive models.  The <span class="pack">mgcv</span> package, ... wait for it..., also allows one to use a spline form of Gaussian process `bs='gp'`.
