---
title: "Polynomials"
author: |
  | Michael Clark
  | Statistician Lead
  | Consulting for Statistics, Computing and Analytics Research
  | <span style="color:'#00274c'">Advanced Research Computing</span>
date: '`r Sys.Date()`'
output:
  html_document:
    css: workshop.css
    highlight: tango
    theme: cosmo
    toc: yes
  pdf_document:
    highlight: pygments
    toc: yes
always_allow_html: yes
---
\pagebreak

# The Polynomial Approach

$$y = f(x) + e$$
$$f(x) = sin(2(4x-2)) + 2e^{-(16^2)(x-.5)^2}$$
$$e \sim N(0,.3^2)$$

```{r simData}
set.seed(123)
x = runif(500)
mu = sin(2*(4*x-2)) + 2*exp(-(16^2)*((x-.5)^2))
y = rnorm(500, mu, .3)
d = data.frame(x,y) 
```

Let's take a look at it visually.

```{r simDataPlot, echo=F}
library(rbokeh)
figure(width=700, theme=bk_ggplot_theme, tools=NULL) %>% 
  ly_points(x, y,line_alpha=0, color='black', size=8)
```

## Polynomial regression doesn't work
In the past people would try and use polynomial regression, e.g. fitting a quadratic or cubic function within the standard regression framework.  However, this is unrealistic at best and at worst isn't useful for complex relationships. In the following even with a polynomial of degree 15 the fit is fairly poor in many areas.

```{r polyreg, echo=FALSE}
fits = sapply(seq(3,15, 3), function(p) fitted(lm(y~poly(x,p)))) %>% 
  data.frame() %>% 
  cbind(x, y, .) %>% 
  gather(key=polynomial, value=fits, -x, -y) %>% 
  mutate(polynomial = factor(polynomial, labels=seq(3,15, 3)))


# fits %>% 
#   ggvis(~x, ~y) %>% 
#   layer_points(fillOpacity:=.5, data=d) %>% 
#   group_by(polynomial) %>% 
#   layer_lines(~x, ~fits, stroke=~polynomial, strokeWidth:=3)

figure(data=fits %>% arrange(x), width=700, theme=bk_ggplot_theme, tools=NULL) %>% 
  ly_points(x, y, color='black', alpha=.5, size=8, line_alpha=0, data=d) %>% 
  ly_lines(x, fits, color=polynomial, width=3) %>% 
  theme_legend(border_line_alpha=0, glyph_width=20, 
               label_standoff=5, legend_padding=3,
               label_text_font_size='8pt')
```

## Piecewise polynomial
We can divide the data into chunks at various points (<span class='emph'>knots</span>), and fit a polynomial within that subset of data.

```{r piecewisePoly, echo=FALSE}
knots = seq(0,1, by=.1)
d$xcut = cut(x, knots, right=F)
d$xcut = factor(d$xcut, levels=c('Int', levels(d$xcut)))
da = arrange(d, xcut, x)
# d$fits = predict(lmList(y~poly(x, 3)|xcut, d))
fits = sapply(levels(da$xcut)[-1], function(g) fitted(lm(y~poly(x, 3), data=filter(da, xcut==g))), simplify=F)

for (i in seq_along(fits)){
  names(fits[[i]]) = NULL
}
xs = sapply(levels(d$xcut)[-1], function(g) da[da$xcut==g, 'x'])
names(fits) = names(xs) = NULL

# d %>%
#   ggvis(~x, ~y) %>%
#   layer_points(fillOpacity:=.5) %>%
#   group_by(xcut) %>%
#   layer_model_predictions(model='lm', formula=y~poly(x,3), strokeWidth:=3, stroke:='darkred')

# lines apparently can't handle noncontinuous
# figure(d) %>% 
#   ly_points(x, y, group=xcut, color='black') %>% 
#   ly_lines(x, fits, color='darkred', width=3)

figure(d, width=700, theme=bk_ggplot_theme, tools=NULL) %>%
  ly_points(x, y, group=xcut, color='black', size=8, line_alpha=0, alpha=.5) %>%
  ly_multi_line(xs=xs, ys=fits, color='darkred', width=3)
```

While this is better, again it is unsatisfactory. The separate fits are unconnected, leading to sometimes notably different predictions for values close together.

## Polynomial spline
Polynomial splines seek to remedy the above problems by creating a smooth function while still taking this piecewise type of approach.  To demonstrate, we'll create a polynomial spline by hand. The following is based on Fahrmeier et al. (2013).

The approach is defined as follows, with $\kappa$ knots on the interval $[a,b]$ as $a=\kappa_1 < ... < \kappa_m =b$:

$$y_i = \gamma_1 + \gamma_2X_i + ... + \gamma_{l+1}(X_i)_+^l + \gamma_{l+2}(X_i - \kappa_2)_+^l ... + \gamma_{l+m-1}(X_i - \kappa_{m-1})_+^l + e_i$$


$$(X_i - \kappa_j)^l_+= \begin{cases} 
(X_i-\kappa_j)^l & X_i \geq k_j \\ 
0 & \textrm{otherwise}
\end{cases}$$

Note that there is nothing special about this. It is just a standard linear regression model when everything is said and done. The first part is even a standard polynomial of degree $l$.

More generally and cleanly:
$$y = f(X) + e = \sum_{j=1}^{d}B_j(X)\gamma_j + e$$

Above, $B$ is a basis function that transforms $X$ as needed, and the $\gamma$ are the regression coefficients.  But before getting too far let's see it in action. Here our polynomial spline will be done with degree $l$ equal to 1, which means that we are just fitting a linear regression.

```{r csbs}
knots = seq(0, 1, by=.1); knots = knots[-length(knots)]  # don't need the last value
l = 1
bs = sapply(1:length(knots), function(k) ifelse(x >= knots[k], (x-knots[k])^l, 0))

head(bs)
```

If we plot this against our target variable $y$, it doesn't look like much.

```{r csbsPlot, echo=FALSE}
library(tidyr)
bs = data.frame(int=1, bs)
d2 = data.frame(x, bs) %>% 
  gather(key=bs, value=bsfunc, -x)


# d2 %>% 
#   ggvis(~x, ~bsfunc) %>% 
#   layer_points(~x, ~y, fillOpacity:=.5, data=d) %>%
#   group_by(bs) %>%
#   layer_lines(stroke =~ bs) %>% 
#   lazerhawk::ggClean()

figure(d2, width=700, theme=bk_ggplot_theme, tools=NULL) %>%
  ly_points(x, y, color='black', size=8, alpha=.5, line_alpha=0, data=d) %>% 
  ly_lines(x, bsfunc, color=bs, data=arrange(d2, x), legend=F)
```

If we multiply each basis by it's corresponding regression coefficient we can start to interpret the result. 

```{r csbsScaledPlot, echo=c(2:4,6:7), eval=-1}
# bs = as.matrix(data.frame(int=1, bs))
lmMod = lm(y~.-1, bs)
bscoefs = coef(lmMod)
bsScaled = sweep(bs, 2, bscoefs,`*`)
colnames(bsScaled) = c('int', paste0('X', 1:10))
head(bsScaled)
bscoefs

d3 = data.frame(x, bsScaled) %>% 
  gather(key=bs, value=bsfunc, -x) %>% 
  dplyr::filter(bsfunc >= min(y) & bsfunc <= max(y))

# d3 %>% 
#   ggvis(~x, ~bsfunc) %>% 
#   layer_points(~x, ~y, fillOpacity:=.5, data=d) %>% 
#   group_by(bs) %>%
#   layer_lines(stroke =~ bs, strokeWidth:=2) %>% 
#   add_axis('y', title='y')

figure(d3, width=700, theme=bk_ggplot_theme, tools=NULL) %>%
  ly_points(x, y,color='black', size=8, alpha=.5, line_alpha=0, data=d) %>% 
  ly_lines(x, bsfunc, color=bs, data=arrange(d3, x), width=2) %>% 
  theme_legend(border_line_alpha=0, label_width=5,
               label_standoff=5, legend_padding=1,
               label_text_font_size='6pt')
```

In the plot above, the horizontal blue line represents the global constant ($\gamma_1$, i.e. our intercept, in our formula above). We have a decreasing function starting from that point onward (yellow line). Between .1 and .2 (red line), the coefficient is negative, furthering the already decreasing slope (i.e. steeper downward). The fourth coefficient is positive, which means that between .2 and .3 our decreasing trend is lessening.  Thus our coefficients $j$ tell us the change in slope that occurs at knot $\kappa_{j-1}$.  If this gets you to thinking about interactions in more common model settings, you're on the right track (e.g. adding a quadratic term is just letting x have an interaction with itself; same thing is going on here).

Finally if plot the sum of the basis functions, which is the same as taking our fitted values from a regression on the basis expansion of X, we get the following fit to the data. And we can see the trend our previous plot suggested.

```{r csbsFitPlotDegree1, echo=FALSE}
plotData = rbind(data.frame(d, sum=fitted(lmMod)),
                 data.frame(x=0, y=bscoefs[1], sum=NA, xcut='Int'))
# plotData %>% na.omit %>% 
#   ggvis(~x, ~y) %>% 
#   layer_points(fillOpacity:=.5, fill=~xcut, data=plotData) %>% 
#   layer_lines(y=~sum, strokeWidth:=2) %>% 
#   group_by(xcut) %>% 
#   layer_lines(y=~sum, stroke=~xcut, strokeWidth:=2) %>% 
#   add_axis('y', title='y')

filter(plotData, xcut!='Int') %>% droplevels %>% 
  figure( width=700, theme=bk_ggplot_theme, tools=NULL) %>%
  ly_points(x, y, color=xcut, size=8, line_alpha=0) %>% 
  ly_lines(x, y=sum, color=xcut, legend=F) %>% 
  ly_points(x, y, color=xcut, size=20, data=filter(plotData, xcut=='Int'))%>% 
  theme_legend(border_line_alpha=0, label_width=5, label_height=2,
               label_standoff=5, legend_spacing=0, legend_padding=0,
               label_text_font_size='8pt')
```


One of the most common approaches uses a cubic spline fit. So we'll change our polynomial degree to 3.  Seems to be fitting the data well.

```{r csFitPlotDegree3, echo=FALSE}
l = 3
bs = sapply(1:length(knots), function(k) ifelse(x >= knots[k], (x-knots[k])^l, 0))
lmModCubicSpline = lm(y~poly(x,3) + bs)
# data.frame(x, y, fits=fitted(lmModCubicSpline)) %>% 
#   ggvis(~x, ~y) %>% 
#   layer_points(~x, ~y, fillOpacity:=.5) %>% 
#   layer_lines(y=~fits, stroke:='ff5503', strokeWidth:=2) %>% 
#   add_axis('y', title='y')

data.frame(x, y, fits=fitted(lmModCubicSpline)) %>% arrange(x) %>% 
  figure( width=700, theme=bk_ggplot_theme, tools=NULL) %>%
  ly_points(x, y, color='black', size=8, line_alpha=0, alpha=.5) %>% 
  ly_lines(x, y=fits, color='#ff5504', width=2)
```

Let's compare it to the <span class='func'>gam</span> function from the <span class='pack'>mgcv</span> package. As we'll see later we won't usually specify the knots directly, and even as we have set things up similar very close to our own approach, the gam function is still doing some things our by-hand approach is not (penalized regression).  We still get pretty close agreement.

```{r csFitPlotvsGAM, echo=FALSE}
library(mgcv)
# data.frame(x, y, fits=fitted(lmModCubicSpline)) %>% 
#   ggvis(~x, ~y) %>% 
#   layer_points(~x, ~y, fillOpacity:=.5) %>% 
#   layer_lines(y=~fits, stroke:='ff5503', strokeWidth:=3) %>% 
#   layer_model_predictions(model='gam', formula=y~s(x, bs='cr'), model_args=list(knots=list(x=knots)), 
#                           stroke:='red', strokeOpacity:=.5) %>% 
#   add_axis('y', title='y')

data.frame(x, y, fits=fitted(lmModCubicSpline), fitsGam=fitted(gam(y~s(x, bs='cr'), knots=list(x=knots)))) %>% 
  arrange(x) %>% 
  figure( width=700, theme=bk_ggplot_theme, tools=NULL) %>%
  ly_points(x, y, color='black', size=8, line_alpha=0, alpha=.5)  %>% 
  ly_lines(x, y=fits, color='#ff5504', width=2) %>% 
  ly_lines(x, y=fitsGam, color='red', width=2)
```


We can see that we're on the right track by using the contructor function within mgcv and a custom function for truncated power series like what we're using above. See the example in the help file for <span class='func'>smooth.construct</span> for the underlying truncated power series function.  I only show a few of the columns, but our by-hand construction and that used by gam are identical to several decimal places. 

```{r gamDatavsByhandData, echo=-1}
smooth.construct.tr.smooth.spec<-function(object,data,knots)
## a truncated power spline constructor method function
## object$p.order = null space dimension
{ m <- object$p.order[1]
  if (is.na(m)) m <- 3 ## default 
  if (m<1) stop("silly m supplied")
  if (object$bs.dim<0) object$bs.dim <- 10 ## default
  nk<-object$bs.dim-m-1 ## number of knots
  if (nk<=0) stop("k too small for m")
  x <- data[[object$term]]  ## the data
  x.shift <- mean(x) # shift used to enhance stability
  k <- knots[[object$term]] ## will be NULL if none supplied
  if (is.null(k)) # space knots through data
  { n<-length(x)
    k<-quantile(x[2:(n-1)],seq(0,1,length=nk+2))[2:(nk+1)]
  }
  if (length(k)!=nk) # right number of knots?
  stop(paste("there should be ",nk," supplied knots"))
  x <- x - x.shift # basis stabilizing shift
  k <- k - x.shift # knots treated the same!
  X<-matrix(0,length(x),object$bs.dim)
  for (i in 1:(m+1)) X[,i] <- x^(i-1)
  for (i in 1:nk) X[,i+m+1]<-(x-k[i])^m*as.numeric(x>k[i])
  object$X<-X # the finished model matrix
  if (!object$fixed) # create the penalty matrix
  { object$S[[1]]<-diag(c(rep(0,m+1),rep(1,nk)))
  }
  object$rank<-nk  # penalty rank
  object$null.space.dim <- m+1  # dim. of unpenalized space
  ## store "tr" specific stuff ...
  object$knots<-k;object$m<-m;object$x.shift <- x.shift
 
  object$df<-ncol(object$X)     # maximum DoF (if unconstrained)
 
  class(object)<-"tr.smooth"  # Give object a class
  object
}

xs = scale(x, scale=F)
bs = sapply(1:length(knots), function(k) ifelse(x >= knots[k], (x-knots[k])^l, 0))
sm = smoothCon(s(x, bs='tr', k=14), data=d, knots=list(x=knots))[[1]]
head(sm$X[,1:6])
modelMatrix = cbind(1, xs, xs^2, xs^3, bs)
identical(round(sm$X,4), round(modelMatrix,4))
```

## The number of knots and where to put them.

A natural question may arise as to how many knots to use. More knots mean more 'wiggliness', as demonstrated here.


```{r differentNKnots, echo=FALSE}
library(dplyr)
l = 3
nk = c(3, 6, 9, 15)
bs = lmCS = list()
for (i in 1:length(nk)){
  knots = seq(0, 1, length.out=nk[i]); knots = knots[-length(knots)]
  # knots[[i]]
  bs[[i]] = sapply(1:length(knots), function(k) ifelse(x >= knots[k], (x-knots[k])^l, 0))
  lmCS[[i]] = lm(y~poly(x,3) + bs[[i]])
}
fits = sapply(lmCS, fitted)

# data.frame(x, y, fits=fits) %>%
#   gather(key='knots', value='value', -x, -y) %>% 
#   mutate(knots = factor(knots, labels=nk)) %>% 
#   ggvis(~x, ~y) %>% 
#   layer_points(fillOpacity:=.1) %>% 
#   group_by(knots) %>% 
#   layer_lines(y=~value, stroke=~knots, strokeWidth:=2) %>% 
#   lazerhawk::ggClean()

data.frame(x, y, fits=fits) %>% arrange(x) %>% 
  gather(key='knots', value='value', -x, -y) %>%  
  mutate(knots = factor(knots, labels=nk)) %>% 
  figure( width=700, theme=bk_ggplot_theme, tools=NULL) %>%
  ly_points(x, y, color='black', size=8, line_alpha=0, alpha=.1)  %>% 
  ly_lines(x, y=value, color=knots, width=2) %>% 
  theme_legend(border_line_alpha=0, label_width=5, label_height=2,
               label_standoff=5, legend_spacing=0, legend_padding=0,
               label_text_font_size='8pt')
  
```

As an example of other types of smooth.terms we might use, here are the basis functions for b-splines. They work notably differently, e.g. over intervals of $l+2$ knots

```{r bSpline, echo=FALSE}
bfs = splines::bs(x, knots=knots[-1])
bsMod = lm(y ~ bfs)
fits = fitted(bsMod)
bfsScaled = sweep(cbind(Int=1, bfs), 2, coef(bsMod),`*`)

bSplineXmatsc = data.frame(x, fits, bfsScaled) %>%
  gather(key=bs, value=bsfunc, -x, -fits)
bSplineXmat = data.frame (Int=1, x = x, bfs) %>% 
  gather(key=bs, value=bsfunc, -x)



# bSplineXmat %>% 
#   ggvis(~x, ~bsfunc) %>% 
#   layer_points(~x, ~y, fillOpacity:=.1, data=d) %>%
#   group_by(bs) %>%
#   layer_lines(stroke=~bs) %>% 
#   hide_legend('stroke') %>% 
#   lazerhawk::ggClean()
# 
# bSplineXmatsc %>% 
#   ggvis(~x, ~bsfunc) %>% 
#   layer_points(~x, ~y, fillOpacity:=.1, data=d) %>%
#   layer_lines(y=~fits, strokeWidth:=2) %>% 
#   group_by(bs) %>%
#   layer_lines(stroke=~bs) %>% 
#   hide_legend('stroke') %>% 
#   lazerhawk::ggClean()


bSplineXmat %>% arrange(x) %>% 
  figure(width=700, theme=bk_ggplot_theme, tools=NULL) %>%
  ly_points(x, y, color='black', size=8, line_alpha=0, alpha=.5, data=d)  %>% 
  ly_lines(x, bsfunc, color=bs, width=2, legend=F) 
bSplineXmatsc %>% arrange(x) %>% 
  figure(width=700, theme=bk_ggplot_theme, tools=NULL) %>%
  ly_points(x, y, color='black', size=8, line_alpha=0, alpha=.5, data=d)  %>% 
  ly_lines(x, bsfunc, color=bs, width=2, legend=F) 
```



